Using Monte Carlo Algorithms to Achieve Resiliency \& Performance at Scale for Linear and Nonlinear Solver Applications

### 2016-01-25 doc/ecp/pres.tex
Updated first section draft.

\setbeamercolor{item}{fg=MSUgreen}
\beamertemplatenavigationsymbolsempty
\setbeamertemplate{headline}{}
\setbeamertemplate{footline}{}
\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{5.5cm}
      {\RaggedRight \bf \large \textcolor{MSUgreen}{\inserttitle}}\\
      \vspace{\baselineskip}\\
      \insertauthor\\
      \vspace{0.5\baselineskip}\\
      \insertinstitute\\
      \insertdate
    \end{minipage}
  \end{tabular}
}

@@

\begin{frame}
  \tableofcontents[pausesections]
\end{frame}

@@

@@

@@

  \begin{itemize}
  \item {\bf MCLS} (\underline{github.com/sslattery/MCLS})
    \begin{itemize}
    \item Massively parallel (MPI) implementation of domain decomposed MCSA
      algorithm
    \item Implemented in Trilinos framework and callable through Trilinos'
      abstract numerical algorithms (Thyra, Stratimikos) interface
    \item Works with any Tpetra row matrix
    \end{itemize}
  \item {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
    \begin{itemize}
    \item GPU-accelerated on-node MC solve
    \item Domain-local MC solves implemented for distributed systems as an
      additive Schwartz method
    \end{itemize}
  \end{itemize}

>>>>

\setbeamercolor{item}{fg=MSUgreen}
\beamertemplatenavigationsymbolsempty
\setbeamertemplate{headline}{}
\setbeamertemplate{footline}{
  \insertframenumber/\inserttotalframenumber}
\setbeamertemplate{title page}
{
  \begin{tabular}{cr}
    \begin{minipage}{5.5cm}
      {\RaggedRight \bf \large \textcolor{MSUgreen}{\inserttitle}}\\
      \vspace{\baselineskip}\\
      \insertauthor\\
      \vspace{0.5\baselineskip}\\
      \insertinstitute\\
      \insertdate
    \end{minipage}
  \end{tabular}
}

@@

\begin{frame}
  \tableofcontents
\end{frame}

@@

\begin{frame}{Domain Decomposed Monte Carlo}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing not yet addressed
      \end{itemize}
    \end{column}

    \begin{column}{0.5\textwidth}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.75}{ \input{ddnu_example.pdftex_t} }
        \end{center}
        \caption{\small Domain decomposition example illustrating
          how domain-to-domain transport creates communication costs.}
      \end{figure}
    \end{column}
  \end{columns}

@@

\begin{frame}{Asynchronous Monte Carlo Transport Kernel}

  \begin{columns}

    \begin{column}{0.5\textwidth}
      \vspace{-0.1in}
      \small
      \begin{itemize}
      \item General extension of the Milagro algorithm (LANL)
      \item Asynchronous nearest neighbor communication of histories
      \item System-tunable communication parameters of buffer size and check
        frequency (performance impact)
      \item Need an asynchronous strategy for exiting the transport loop
        without a collective (running sum)
      \end{itemize}

      \vspace{-0.15in}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.4}{ \input{domain_to_domain.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

    \begin{column}{0.5\textwidth}

      \vspace{-0.25in}
      \begin{figure}[htpb!]
        \begin{center}
          \scalebox{0.825}{ \input{transport_loop.pdftex_t} }
        \end{center}
      \end{figure}

    \end{column}

  \end{columns}

\begin{frame}{Parallel Test - Simplified $P_N$ ($SP_N$) Assembly Problem}
\small
\begin{columns}

    \begin{column}{0.55\textwidth}

      \vspace{-0.2in}
      \begin{figure}
        \centering
        \includegraphics[width=2.0in]{../siam_cse_2015/presentation/prob4}
        \vspace{-0.2in}
        \caption{$SP_N$ solution example}
      \end{figure}

      \vspace{-0.1in}

      The ($SP_N$) equations are an approximation to the Boltzmann neutron
      transport equation used to simulate nuclear reactors

      \vspace{-0.2in}

      \[
        \small
        -\nabla \cdot \mathbb{D}_n \nabla \mathbb{U}_n + \sum_{m=1}^4
        \mathbb{A}_{nm} \mathbb{U}_m = \frac{1}{k} \sum_{m=1}^4
        \mathbb{F}_{nm} \mathbb{U}_m
      \]

    \end{column}

    \begin{column}{0.45\textwidth}
      Scaling problem -- $1 \times 1$ to $17 \times 17$ array of fuel
      assemblies with 289 pins each resolved by a $2 \times 2$ spatial mesh
      and 200 axial zones

      \vfill

      \begin{itemize}
      \item 7 energy groups, 1 angular moment, 1.6M to 273.5M degrees of
        freedom
      \item 64 to 10,800 computational cores via domain decomposition
      \item We are usually interested in solving generalized eigenvalue
        problem - we use the operator from these problems to test the kernel
        scaling
      \end{itemize}
    \end{column}

  \end{columns}

\begin{frame}{Domain Decomposition Scaling}

  \vspace{-0.2in}

  \begin{table}[htb!]
    \tiny
    \begin{center}
      \begin{tabular}{rrrrrrr}
        \toprule
        \multicolumn{1}{r}{Cores} &
        \multicolumn{1}{r}{DOFs} &
        \multicolumn{1}{r}{DOFs/Core} &
        \multicolumn{1}{r}{Time Min (s)} &
        \multicolumn{1}{r}{Time Max (s)} &
        \multicolumn{1}{r}{Time Ave (s)} &
        \multicolumn{1}{r}{Efficiency}
        \\ \midrule
        %%
        256 & 273\,509\,600 & 1\,068\,397 & 515.51 & 515.52 & 515.52 & 1.00 \\
        1\,024 & 273\,509\,600 & 267\,099 & 122.76 & 122.77 & 122.76 & 1.05 \\ %CACHE
        4\,096 & 273\,509\,600 & 66\,775 & 27.96 & 27.97 & 27.96 & 1.15 \\
        7\,744 & 273\,509\,600 & 35\,319 & 17.72 & 17.72 & 17.72 & 0.96 \\
        10\,816 & 273\,509\,600 & 25\,288 & 13.72 & 13.72 & 13.72 & 0.89 \\
        %%
        \bottomrule
      \end{tabular}
    \end{center}
    \vspace{-0.07in}
    \caption{\small Strong Scaling}
  \end{table}

  \vspace{-0.4in}

  \begin{table}[htb!]
    \tiny
    \begin{center}
      \begin{tabular}{rrrrrrr}
        \toprule
        \multicolumn{1}{r}{Cores} &
        \multicolumn{1}{r}{DOFs} &
        \multicolumn{1}{r}{DOFs/Core} &
        \multicolumn{1}{r}{Time Min (s)} &
        \multicolumn{1}{r}{Time Max (s)} &
        \multicolumn{1}{r}{Time Ave (s)} &
        \multicolumn{1}{r}{Efficiency}
        \\ \midrule
        %%
        64 & 1\,618\,400 & 25\,288 & 12.65 & 12.65 & 12.65 & 1.00 \\
        256 & 6\,473\,600 & 25\,288 & 12.86 & 12.86 & 12.86 & 0.98 \\
        1\,024 & 25\,894\,400 & 25\,288 & 14.59 & 14.59 & 14.59 & 0.87 \\
        4\,096 & 103\,577\,600 & 25\,288 & 14.25 & 14.25 & 14.25 & 0.89 \\
        7\,744 & 195\,826\,400 & 25\,288 & 14.75 & 14.78 & 14.75 & 0.86 \\
        10\,816 & 273\,509\,600 & 25\,288 & 13.72 & 13.72 & 13.72 & 0.92 \\
        %
        \bottomrule
      \end{tabular}
    \end{center}
        \vspace{-0.07in}
    \caption{\small Weak Scaling}
  \end{table}

  \vspace{-0.25in}

  \begin{itemize}
    \small
  \item Titan Cray XK7 was used with no GPUs - used Eos parameters
  \item Results are consistent with transport literature
  \item History length of 15 and 3 histories per DOF
  \item Consider these effective operator apply times of 15 realizations of
    $\ve{H}$
  \item MC dominates run times in solver applications
  \end{itemize}

\begin{frame}{Replication}

  \vspace{-0.1in}

  Different batches of Monte Carlo samples can be combined in summation via
  superposition if they have different random number streams.  Consider each
  of these batches independent \textit{subsets} of a Monte Carlo operator
  where now the operator can be formed as a general additive decomposition of
  $N_S$ subsets:

  \[
  \mathbf{M_{MC}} = \frac{1}{N_S}\sum_{n=1}^{N_s}{\mathbf{M_n}}
  \]

  We replicate the linear problem and form each subset on a different group of
  parallel processes. Applying the subsets to a vector requires an AllReduce
  to form the sum. Each subset is domain decomposed.
  \vfill
  This provides a mechanism to support \textit{resiliency}; the replicated
  component provides {\bf useful} work, not just redundancy

\begin{frame}{Replication Scaling}

  \begin{table}[htb!]
    \tiny
    \begin{center}
      \begin{tabular}{rrrrrrr}
        \toprule
        \multicolumn{1}{r}{Subsets} &
        \multicolumn{1}{r}{Cores} &
        \multicolumn{1}{r}{DOFs} &
        \multicolumn{1}{r}{Time Min (s)} &
        \multicolumn{1}{r}{Time Max (s)} &
        \multicolumn{1}{r}{Time Ave (s)} &
        \multicolumn{1}{r}{Efficiency}
        \\ \midrule
        %%
        1 & 256 & 6\,473\,600 & 12.65 & 12.65 & 12.65 & 1.00 \\
        2 & 512 & 6\,473\,600 & 6.62 & 6.80 & 6.72 & 0.93 \\
        3 & 768 & 6\,473\,600 & 4.50 & 4.73 & 4.66 & 0.89 \\
        4 & 1\,024 & 6\,473\,600 & 3.62 & 3.81 & 3.71 & 0.83 \\
        %
        \bottomrule
      \end{tabular}
    \end{center}
    \caption{\small Fixed algorithmic scaling}
  \end{table}

  \vfill

  \begin{itemize}
    \small
  \item Timings are Monte Carlo time plus AllReduce time
  \item 256 cores per subset with domain decomposition
  \item $1/N_S$ samples per subset gives flat algorithmic performance
  \item Two performance issues
    \begin{itemize}
    \item Dividing up samples creates work starvation
    \item AllReduce buffer size is constant regardless of work starvation
    \end{itemize}
  \item Research is needed for a more performant subset combination
  \end{itemize}

\begin{frame}{GPU Optimization}

  We have implemented a version of MCSA that uses GPUs to perform the MC
  linear solve.

  Distributed parallelism is enabled through an additive Schwartz method, all
  MC solves
  \begin{itemize}
  \item All MC random walks are local
  \item No inter-domain communication of histories
  \end{itemize}

  \begin{table}[h]
    \begin{center}
      \caption{2D parabolic problem with AINV preconditioning.}
      \begin{tabular}{|c|c|c|c|}\hline
        Node & Iterations & Time (sec) & Relative Residual \\\hline
        CPU & 6 & 992.23 & $1.23\times 10^{-7}$ \\\hline
        GPU & 6 & 29.83 & $1.53\times 10^{-7}$ \\\hline
      \end{tabular}
    \end{center}
  \end{table}

%%---------------------------------------------------------------------------%%
\subsection{Research Requirements}
%%---------------------------------------------------------------------------%%

\begin{frame}

  The following important questions need to be answered before MC methods can
  be applied to linear systems as a ``standard'' solver:

  \begin{itemize}
  \item Automatically determine history termination criteria
    \begin{itemize}
    \item We have investigated adaptive schemes based on variance
    \item However, variance is not directly correlated with error
    \end{itemize}
  \item Identify classes of matrices and preconditioners that satisfy
    convergence constraints
    \begin{itemize}
    \item This has been done for SDD and GDD
    \end{itemize}
  \item Demonstrate resiliency to hard MPI errors through replicated domain
    parallelism
  \item Further investigate non-linear, time-dependent problems
    \begin{itemize}
    \item This was part of the original proposal
    \item Non-linear, sub-cycle time-stepping can result in strongly diagonally
      dominant problems for which MCSA has been shown to be effective
    \end{itemize}
  \end{itemize}
\end{frame}

@@

  {\bf MCLS} (\underline{github.com/sslattery/MCLS})
  \begin{itemize}
  \item Massively parallel (MPI) implementation of domain decomposed MCSA
    algorithm
  \item Implemented in Trilinos framework and callable through Trilinos'
    abstract numerical algorithms (Thyra, Stratimikos) interface
  \item Works with any Tpetra row matrix
  \end{itemize}
  \vfill
  {\bf Alea} (\underline{github.com/ORNL-CEES/Profugus})
  \begin{itemize}
  \item GPU-accelerated on-node MC solve
  \item Domain-local MC solves implemented for distributed systems as an
    additive Schwartz method
  \end{itemize}
